Clone and graft:\\ Testing scientific applications as they are built

### 2014-12-03 turcksin-spring-mechanics/paper.tex
Finish writing the text, add bibliography.

Our group has been building scientific software for more than a decade and a
half by now. Over the years we have developed many procedures that help us
test our software extensively but our expertise was mostly in the development
of \textit{libraries} for numerical methods -- specifically, the \dealii{}
library for finite element computations (see
\url{http://www.dealii.org/}). Libraries are of course the foundation of
almost every single scientific code, starting with ``simple'' ones such as
BLAS and LAPACK to very much more complex ones such as \petsc{}, \trilinos{}, or
\dealii{}.

@@

\begin{wrapfigure}{R}{0.44\textwidth}
  \begin{center}
    \vspace*{-24pt}
    \includegraphics[width=0.42\textwidth,height=0.42\textwidth]{figures/aspect.png}
    \vspace*{-12pt}
  \end{center}
  \caption{\it Snapshot of the temperature field of an \aspect{} simulation
    showing convection in the Earth mantle.}
  \vspace*{-3mm}
  \label{fig:aspect}
\end{wrapfigure}
The code that made us think about how to test whole applications is called
\aspect{} -- short for the Advanced Solver for Problems in Earth ConvecTion
\cite{...}. It is a program that simulates how material moves around the Earth
mantle (i.e., the region between the metallic core of Earth and the plates at
the surface on which we live). While the material in the mantle is solid rock,
it is hot and under enormous pressure and can deform with velocities of a few
centimeters per year. On time scales of millions of years, it therefore
behaves like a fluid. Since it is heated from below and cooled from above, it
shows the same kind of behavior as a pot on the stove: blobs of hot material
rise up, cool at the surface, and blobs of cool material fall down. In other
words, it convects. The details of trying to simulate this on a computer go
beyond what we want to discuss here, but it is worth showing a picture and
linking to the videos at \url{http://www.youtube.com/embed/j63MkEc0RRw} and
\url{http://www.youtube.com/embed/EJJ6f4hmDPU} simply because they are pretty.

@@

Over the past 15 years, we have played with many ideas how to address this
problem. For example, we tried to just never output numbers to their full 16
digits of double precision. The idea is that if you only output 5 digits, then
roundoff surely can't play a role. The problem is that the numbers 1.000\,049
999\,999\,999\,9 and 1.000\,050\,000\,000\,000\,0 only differ by round-off,
but when rounded to 5 digits, you either get 1.000\,0 or 1.000\,1. One could
think that this is a contrived example, but it turns out that this happens
with rather frustrating regularity if you just output enough numbers -- as one
often does in scientific codes.

@@

The solution to these problems that we have ultimately come to use
is a program called \texttt{numdiff}. \texttt{numdiff} works essentially like
\texttt{diff} but it recognizes if the text it compares in two files is
actually a number. If two numbers differ by less than some relative or
absolute tolerance, then they are considered equal even if the characters by
which they are represented in an output file are different. For example, you
would replace the \texttt{diff} command used previously by the
following to ensure that numbers $a$ and $b$ compare as equal if $|a-b|\le
10^{-6}$ or $|a-b|\le 10^{-8} \max\{|a|,|b|\}$:
\begin{lstlisting}[frame=single,basicstyle=\footnotesize,language=csh]
  numdiff -a 1e-6 -r 1e-8 testcase-1.output testcase-1.output.new
\end{lstlisting}

\section{Conclusions}
\label{sec:conclusions}

>>>>

Our group has been building scientific software for more than a decade and a
half by now. Over the years we have developed many procedures that help us
test our software extensively but our expertise was mostly in the development
of \textit{libraries} for numerical methods -- specifically, the \dealii{}
library for finite element computations (see
\url{http://www.dealii.org/} and \cite{BHK07,BK99m}). Libraries are of course the foundation of
almost every single scientific code, starting with ``simple'' ones such as
BLAS and LAPACK to very much more complex ones such as \petsc{}, \trilinos{}, or
\dealii{} \cite{petsc,trilinos,trilinos-web-page}.

@@

\begin{wrapfigure}{R}{0.44\textwidth}
  \begin{center}
    \vspace*{-24pt}
    \includegraphics[width=0.42\textwidth,height=0.42\textwidth]{figures/aspect.png}
    \vspace*{-12pt}
  \end{center}
  \caption{\it Snapshot of the temperature field of an \aspect{} simulation
    showing convection in the Earth mantle.}
  \vspace*{-3mm}
  \label{fig:aspect}
\end{wrapfigure}
The code that made us think about how to test whole applications is called
\aspect{} -- short for the Advanced Solver for Problems in Earth ConvecTion
\cite{KHB12}. It is a program that simulates how material moves around the Earth
mantle (i.e., the region between the metallic core of Earth and the plates at
the surface on which we live). While the material in the mantle is solid rock,
it is hot and under enormous pressure and can deform with velocities of a few
centimeters per year. On time scales of millions of years, it therefore
behaves like a fluid. Since it is heated from below and cooled from above, it
shows the same kind of behavior as a pot on the stove: blobs of hot material
rise up, cool at the surface, and blobs of cool material fall down. In other
words, it convects. The details of trying to simulate this on a computer go
beyond what we want to discuss here, but it is worth showing a picture and
linking to the videos at \url{http://www.youtube.com/embed/j63MkEc0RRw} and
\url{http://www.youtube.com/embed/EJJ6f4hmDPU} simply because they are pretty.

@@

Over the past 15 years, we have played with many ideas how to address this
problem. For example, we tried to just never output numbers to their full 16
digits of double precision. The idea is that if you only output 5 digits, then
roundoff surely can't play a role. The problem is that the numbers 1.234\,549
999\,999\,999\,9 and 1.234\,550\,000\,000\,000\,1 only differ by round-off,
but when rounded to 5 digits, you either get 1.234\,5 or 1.234\,6. One could
think that this is a contrived example, but it turns out that this happens
with rather frustrating regularity if you just output enough numbers -- as one
often does in scientific codes.

@@

The solution to these problems that we have ultimately come to use
is a program called \texttt{numdiff}. \texttt{numdiff} works essentially like
\texttt{diff} but it recognizes if the text it compares in two files is
actually a number. If two numbers differ by less than some relative or
absolute tolerance, then they are considered equal even if the characters by
which they are represented in an output file are different. For example, you
would replace the \texttt{diff} command used previously by the
following to ensure that numbers $a$ and $b$ compare as equal if $|a-b|\le
10^{-6}$ or $|a-b|\le 10^{-8} \min\{|a|,|b|\}$:
\begin{lstlisting}[frame=single,basicstyle=\footnotesize,language=csh]
  numdiff -a 1e-6 -r 1e-8 testcase-1.output testcase-1.output.new
\end{lstlisting}
Using this trick, numbers that are written into the output file
\texttt{testcase-1.output.new} on one machine with slight differences to the
ones originally saved in \texttt{testcase-1.output} will still be considered
the same, and the comparison will succeed.

\section{Summary}
\label{sec:conclusions}

When we talk to students who are just starting writing their own codes, then
they typically understand the importance of testing in some abstract
sense. But almost all of them have two fundamental misconceptions: First, that
their codes are either written well enough or will not be changed in ways
fundamental enough to suggest that these changes might break existing
functionality. And second, that doing adequate testing is just too hard and
cumbersome to be a justifiable expense of time.

The first of these two points is empirically false. Only bad programmers
expect that they will not break existing functionality, no matter how trivial
a change might be. All experienced programmers will know this to be wrong. The
reason is that over the course of a scientific code -- say, over the course of
the 3 years a graduate student does research, and maybe even beyond this if
the code is handed on to the next student, or if you go on to be a postdoc --
a code starts out small but is constantly developed into something much larger
with much more functionality in all sorts of directions. You're not going to
create one essential version of the code that you can verify to be correct and
then just round off some edges; you're in fact going to replace the linear by
a nonlinear model, replace the direct solver by an iterative one, add terms to
the equation, and make other very fundamental changes that will require you to
edit all throughout your code. There is no other way than systematic testing
to ensure that the computation you ran today will still run tomorrow. And you
will have to run it again tomorrow: your adviser or boss may ask you to redo a
computation with some slight changes, a reviewer of a paper might ask to do
additional simulations, you may have forgotten the exact parameters for a data
set you want to use in your thesis and will have to redo it to verify,
etc. You may think that once you have created the data 
for some particular simulation, that the issue is done and will not have to
repeated -- i.e., that you might in fact just throw away the code -- but this
is not typically true.

The second of the poinst above -- that testing is hard and time consuming --
is also empirically false. Once you start to do it systematically, all you
have to do is morph an existing input file into a new one that tests the
aspect you have just implemented, and add it to your testsuite. It doesn't
take very long, you just have to do it right from the start. Once you do this,
you will find how much faster it suddenly becomes to debug codes. To us, one
of the most surprising facts of developing stringent testing procedures is not
just how much better our codes become, but in fact how much safer we feel
making changes in central places of our code: without tests, we always felt
insecure and wondered what a change might do to the code, whether it is
correct in all cases, and if we might be forgetting some cases. With tests,
we find ourselves a lot more at ease: just make a change, run the 140 or so
tests and see what happens -- if it all continues to work, then that may not
be a guarantee that a change is correct, but it sure makes you sleep better at
night. It really isn't all that hard to find peace of mind!

\bibliographystyle{alpha}
\bibliography{paper}
